---
title: "C.3 Differential Expression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{C.3 Differential Expression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  ## too much memory for GitHub actions
  eval = !nzchar(Sys.getenv("BIOCONDUCTOR_DOCKER_VERSION"))
)
```

# Orchestrating Single-Cell Analysis with _Bioconductor_

[OSCA][] -- an amazing resource!

[OSCA]: https://bioconductor.org/books/OSCA

# Differential expression

This script is derived from the [OSCA Multi-Sample][OSCA-M] book,
[Chapter 4][OSCA-M-4]: DE analyses between conditions. See the book
for full details.

```{r, echo = FALSE, message = FALSE}
library(MouseGastrulationData)
library(scater)
library(scran)
library(batchelor)
requireNamespace("igraph")
```

## Initial analysis

```{r, message = FALSE}
## load data
library(MouseGastrulationData)
sce.chimera <- WTChimeraData(samples=5:10)
sce.chimera

## feature annotation
library(scater)
rownames(sce.chimera) <- uniquifyFeatureNames(
    rowData(sce.chimera)$ENSEMBL, rowData(sce.chimera)$SYMBOL
)

## qc
drop <- sce.chimera$celltype.mapped %in% c("stripped", "Doublet")
sce.chimera <- sce.chimera[,!drop]

## normalization
sce.chimera <- logNormCounts(sce.chimera)

## variance modeling
library(scran)
dec.chimera <- modelGeneVar(sce.chimera, block=sce.chimera$sample)
chosen.hvgs <- dec.chimera$bio > 0

## merge batches
library(batchelor)
set.seed(01001001)
merged <- correctExperiments(
    sce.chimera, 
    batch=sce.chimera$sample, 
    subset.row=chosen.hvgs,
    PARAM=FastMnnParam(
        merge.order=list(
            list(1,3,5), # WT (3 replicates)
            list(2,4,6)  # td-Tomato (3 replicates)
        )
    )
)

## cluster
g <- buildSNNGraph(merged, use.dimred="corrected")
clusters <- igraph::cluster_louvain(g)
colLabels(merged) <- factor(clusters$membership)

## dimensionality reduction
merged <- runTSNE(merged, dimred="corrected", external_neighbors=TRUE)
merged <- runUMAP(merged, dimred="corrected", external_neighbors=TRUE)
```


## Differential expression

Initial observations -- labels are similar, whether a 'td-Tomato'
positive cell or not.

```{r}
colData(merged) |>
    dplyr::as_tibble() |>
    dplyr::count(label, tomato) |>
    tidyr::pivot_wider(names_from = "tomato", values_from = "n")

plotTSNE(merged, colour_by="tomato", text_by="label", point_size = .1)
```

Create pseudo-bulk samples

```{r}
# Using 'label' and 'sample' as our two factors; each column of the output
# corresponds to one unique combination of these two factors.
summed <- aggregateAcrossCells(
    merged, 
    id=colData(merged)[, c("celltype.mapped", "sample")]
)
summed
```

[OSCA-M]: https://bioconductor.org/books/OSCA.multisample/
[OSCA-M-4]: https://bioconductor.org/books/release/OSCA.multisample/multi-sample-comparisons.html

# Session information

This document was produced with the following *R* software:

```{r session_info}
sessionInfo()
```

