---
title: "C. R / Bioconductor for Genomic Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{C. R / Bioconductor for Genomic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This workshop walks through a single-cell RNAseq analysis using
[Seurat][] and *R* / *Bioconductor* packages in the online resource
[Orchestrating Single-Cell Analysis with Bioconductor][OSCA]. The
workshop assumes some familiarity with *R*, and sufficient domain
knowledge to know at a superficial level the technical details and
scientific motivation for single-cell analysis. The main goal is to
increase participants' confidence in using *R* to embark on creative
and critical data exploration -- analysis of single cell data is
seldom straight-forward, requiring understanding of methods and
critical assessment of data at many steps.

[OSCA]: https://bioconductor.org/books/release/OSCA/
[Seurat]: https://satijalab.org/seurat/index.html

# Session information

This document was produced with the following *R* software:

```{r session_info}
sessionInfo()
```

