---
title: "F. Annotating cell types in Bioconductor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{F. Annotating cell types in Bioconductor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Orchestrating Single-Cell Analysis with _Bioconductor_

[OSCA][] -- an amazing resource!

[OSCA]: https://bioconductor.org/books/OSCA

# Annotating cell types

This script is derived from the [OSCA Basic][OSCA-B] book, [Chapter
7][OSCA-B-7]: Cell type annotation. See the book for full details.

```{r, echo = FALSE, message = FALSE}
library(DropletTestFiles)
library(DropletUtils)
library(scater)
library(EnsDb.Hsapiens.v86)
library(scran)
library(celldex)
library(SingleR)
```

## Initial analysis

```{r}
## data retrieval
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)

## input
library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)

## annotation
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol
)

## qc
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[, which(e.out$FDR <= 0.001)]

## normalization
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)

## variance modeling
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)

## dimensionality reduction
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")

## clustering
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)
```

## Using curated reference data

Use a curated reference data set, in this case derived from Blueprint
and ENCODE data.

```{r}
library(celldex)
ref <- BlueprintEncodeData()
colData(ref) |>
    dplyr::as_tibble()
colData(ref) |>
    dplyr::as_tibble() |>
    dplyr::count(label.main, sort = TRUE)
```

Annotate our data using the 'main' label for cell type

```{r, warning = FALSE}
library(SingleR)
## assign each cell in sce.pbmc to a type
pred <- SingleR(test=sce.pbmc, ref=ref, labels=ref$label.main)
pred |>
    dplyr::as_tibble(rownames = "cell_id")
pred |>
    dplyr::as_tibble() |>
    dplyr::count(labels)

plotScoreHeatmap(pred)
```

## Other strategies

Other strategies ouline in [Chapter 7][OSCA-B-7]

- Custom references
- Gene sets
- Markers

[OSCA-B]: https://bioconductor.org/books/OSCA.basic
[OSCA-B-7]: https://bioconductor.org/books/release/OSCA.basic/cell-type-annotation.html

# Session information

This document was produced with the following *R* software:

```{r session_info}
sessionInfo()
```

```{r include = FALSE}
## clean up .GlobalEnv at end of vignette
rm(list = ls(envir = .GlobalEnv, all.names = TRUE), envir = .GlobalEnv)
```
