---
title: "F. Annotating cell types in Bioconductor"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{F. Annotating cell types in Bioconductor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Orchestrating Single-Cell Analysis with _Bioconductor_

[OSCA][] -- an amazing resource!

[OSCA]: https://bioconductor.org/books/OSCA

# Annotating cell types

This script is derived from the [OSCA Basic][OSCA-B] book, [Chapter
7][OSCA-B-7]: Cell type annotation. See the book for full details.

```{r, echo = FALSE, message = FALSE}
library(DropletTestFiles)
library(DropletUtils)
library(scater)
library(EnsDb.Hsapiens.v86)
library(scran)
library(celldex)
library(SingleR)
```

## Initial analysis

We start by retrieving some sample DropSeq data.

```{r}
library(DropletTestFiles)
raw.path <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/raw.tar.gz")
out.path <- file.path(tempdir(), "pbmc4k")
untar(raw.path, exdir=out.path)
```

Input the data using DropSeqUtils

```{r}
library(DropletUtils)
fname <- file.path(out.path, "raw_gene_bc_matrices/GRCh38")
sce.pbmc <- read10xCounts(fname, col.names=TRUE)
```

We lightly 'annotate' the data by making sure that feature names are
unique

```{r}
library(scater)
rownames(sce.pbmc) <- uniquifyFeatureNames(
    rowData(sce.pbmc)$ID, rowData(sce.pbmc)$Symbol
)
```

As a quality control step, check for and remove empty droplets.

```{r}
set.seed(100)
e.out <- emptyDrops(counts(sce.pbmc))
sce.pbmc <- sce.pbmc[, which(e.out$FDR <= 0.001)]
```

Normalize cell counts by clustering cells and computing scale factors
per cluster. Update the dataset to include log counts

```{r}
library(scran)
set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
sce.pbmc <- logNormCounts(sce.pbmc)
```

Variance modeling fits a statistical model to the count data; this is
used to identify 'highly variable' genes.

```{r}
set.seed(1001)
dec.pbmc <- modelGeneVarByPoisson(sce.pbmc)
top.pbmc <- getTopHVGs(dec.pbmc, prop=0.1)
```

Calculate reduced-dimensionality representations of the
data. `denoisePCA()` removes technical noise from log-normalized
counts. Technical noise, in contrast to 'biological' signal, is
variation between cells that is uncorrelated across genes. Removing
technical noise improves resolution in subsequent dimensionality
reduction steps. tSNE and UMAP are two common methods for reducing the
genes x samples data to two or a few dimensions for visual
exploration. Generally, UMAP is less sensitive to parameter choice and
random number seed.

```{r}
set.seed(10000)
sce.pbmc <- denoisePCA(sce.pbmc, subset.row=top.pbmc, technical=dec.pbmc)

set.seed(100000)
sce.pbmc <- runTSNE(sce.pbmc, dimred="PCA")

set.seed(1000000)
sce.pbmc <- runUMAP(sce.pbmc, dimred="PCA")
```

Perform shared nearest-neighbor clustering (`buildSNNGraph()`). These
nearest neighbors are then used in community detection algorithms
(`cluster_walktrap()`) to find similar cells. This represents a faster
approach than more traditional hierarchical clustering.

```{r}
g <- buildSNNGraph(sce.pbmc, k=10, use.dimred = 'PCA')
clust <- igraph::cluster_walktrap(g)$membership
colLabels(sce.pbmc) <- factor(clust)
```

## Using curated reference data

One approach to cell annotation is to compare expression profiles to a
curated reference data set. We use the reference data derived from
Blueprint and ENCODE data; see `browseVignettes("celldex")` for a
description of this and other reference datasets.

```{r}
ref <- celldex::BlueprintEncodeData()
colData(ref) |>
    dplyr::as_tibble()
colData(ref) |>
    dplyr::as_tibble() |>
    dplyr::count(label.main, sort = TRUE)
```

Annotate our data using the 'main' label for cell type

```{r, warning = FALSE}
library(SingleR)
## assign each cell in sce.pbmc to a type
pred <- SingleR(test=sce.pbmc, ref=ref, labels=ref$label.main)
pred |>
    dplyr::as_tibble(rownames = "cell_id")
pred |>
    dplyr::as_tibble() |>
    dplyr::count(labels)
```

Finaly, visualize the cell annotation. Generally, B-cells and
Monocytes are well-annotated, e.g., cells labelled 'B-cells' have a
strong (yellow) signal only under the B-cell label; CD4+ and CD8+
cells less well-anotated. It is interesting to explore biological
reasons that might contribute to this weaker classification.

```{r}
plotScoreHeatmap(pred)
```

## Other strategies

Other strategies ouline in [Chapter 7][OSCA-B-7]

- Custom references
- Gene sets
- Markers

[OSCA-B]: https://bioconductor.org/books/OSCA.basic
[OSCA-B-7]: https://bioconductor.org/books/release/OSCA.basic/cell-type-annotation.html

# Session information

This document was produced with the following *R* software:

```{r session_info}
sessionInfo()
```

```{r include = FALSE}
## clean up .GlobalEnv at end of vignette
rm(list = ls(envir = .GlobalEnv, all.names = TRUE), envir = .GlobalEnv)
```
